---
title: "Spatial Analysis R Continuum"
output: html_document
date: "`r Sys.Date()`"
---

## Further Spatial Analysis to the Data.
#1. Try turning the analysis into one piped sequence of code.
#2. Compare health coverage for males and females in the Northern departments.
#3. Compare the health coverage in another area of Haiti.
#4. Try plotting a bar chart of the health coverage by area.
#5. Combine multiple plots using patchwork.


```{r}
#Install and Load the necessary packages to run the tasks.
library(dplyr)
library(readr)
library(ggplot2)
install.packages("patchwork")
library(patchwork)
```

#Before we run the required tasks, we first upload out raw data to R.
```{r}
haiti_data <- read_csv("~/R/isair_r/data/haiti/haiti-healthsites.csv")
View(haiti_data)
```

#1. Turn analysis into a single piped sequence.
```{r}
haiti_summary <- haiti_data %>%
  select(adm1_en, adm2_en, num_health_facilities, total_population) %>%
  filter(adm1_en %in% c("North", "North-East", "North-West")) %>%
  group_by(adm1_en, adm2_en) %>%
  summarise(
    num_health_facilities = sum(num_health_facilities),  # Sum facilities if grouped
    total_population = first(total_population)  # Take first value since it's already grouped
  ) %>%
  mutate(pop_per_health = total_population / num_health_facilities) %>%
  arrange(desc(pop_per_health))

# Print the first few rows of the summary
print(haiti_summary)

# Extract the top 5 worst-served areas.
top_5_worst <- head(haiti_summary, 5)

#Save the trasformed dataset.
write_csv(haiti_summary, "~/R/isair_r/data/haiti/haiti-healthsites.csv")

#Print results
print(top_5_worst)

```



#2. Comparing the health coverage for males and females in Northern departments.
```{r}
#The data provided does not come with gender categories.
```

#3. Compare the health coverage in another area of Haiti.
```{r}
head(haiti_data)

## Data Processing and Piped Analysis
northern_depts <- haiti_data %>%
  filter(adm1_en == "North")

northwest_depts <- haiti_data %>%
  filter(adm1_en == "North-West")


## Compare Health Coverage in Northern Departments
ggplot(northern_depts, aes(x = reorder(adm2_en, pop_per_health), y = pop_per_health)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Health Coverage in Northern Departments",
       x = "District",
       y = "Population per Health Facility")


## Compare Health Coverage in North-West Departments
ggplot(northwest_depts, aes(x = reorder(adm2_en, pop_per_health), y = pop_per_health)) +
  geom_bar(stat = "identity", fill = "orange") +
  coord_flip() +
  labs(title = "Health Coverage in North-West Departments",
       x = "District",
       y = "Population per Health Facility")
```

#4. Combine multiple plots using patchwork.
```{r}
north_plot <- ggplot(northern_depts, aes(x = reorder(adm2_en, pop_per_health), y = pop_per_health)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Health Coverage in Northern Departments",
       x = "District",
       y = "Population per Health Facility")

nw_plot <- ggplot(northwest_depts, aes(x = reorder(adm2_en, pop_per_health), y = pop_per_health)) +
  geom_bar(stat = "identity", fill = "orange") +
  coord_flip() +
  labs(title = "Health Coverage in North-West Departments",
       x = "District",
       y = "Population per Health Facility")

north_plot + nw_plot
```

#Overall Health Coverage by Department
```{r}
ggplot(haiti_data, aes(x = adm1_en, y = pop_per_health, fill = adm1_en)) +
  geom_boxplot() +
  labs(title = "Overall Health Coverage Comparison by Department",
       x = "Department",
       y = "Population per Health Facility") +
  theme(legend.position = "none")
```

